%% ============================================================
%% DeIR-Dual: Dual-Query Instruction Rewriting for Training-Free
%% Instruction-Following Retrieval
%%
%% Target: AAAI 2027
%% Format: AAAI two-column LaTeX
%% ============================================================

\documentclass[letterpaper]{article}
\usepackage{aaai26}
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\usepackage{ifxetex}
\ifxetex\else\usepackage[T1]{fontenc}\fi
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage{enumitem}
\usepackage{xcolor}

% ============================================================
% Custom macros
% ============================================================
\newcommand{\method}{DeIR-Dual}
\newcommand{\qbase}{Q_{\text{base}}}
\newcommand{\qpos}{Q_{\text{pos}}}
\newcommand{\qminus}{Q_{\text{minus}}}
\newcommand{\sbase}{S_{\text{base}}}
\newcommand{\spos}{S_{\text{pos}}}
\newcommand{\sneg}{S_{\text{neg}}}
\newcommand{\sfinal}{S_{\text{final}}}
\newcommand{\todo}[1]{\textcolor{red}{\textbf{[TODO] #1}}}

\hypersetup{pdfinfo={TemplateVersion={2026.1}}}

\setcounter{secnumdepth}{2}

% ============================================================
% Author info (uncomment for final submission)
% ============================================================
% \author{
%   Author One,
%   Author Two,
%   Author Three
% }
% \affiliations{
%   Affiliation 1\\
%   Affiliation 2\\
%   Affiliation 3\\
% }

\begin{document}

% Force write bibstyle to aux (xelatex font errors may skip it)
\makeatletter
\ifx\nobibliographystyle\@undefined
  \write\@auxout{\string\bibstyle{plain}}%
\fi
\makeatother

\title{DeIR-Dual: Dual-Query Instruction Rewriting for Training-Free Instruction-Following Retrieval}
\author{Anonymous Authors}
\date{}
\maketitle

% ============================================================
% ABSTRACT
% ============================================================
\begin{abstract}
Instruction-following retrieval (IF-IR) requires systems to change their
ranking when the query instruction changes.  Existing training-free
query rewriting methods (e.g., HyDE, Query2Doc) improve retrieval
quality through semantic enhancement but largely ignore exclusion
constraints, resulting in near-zero instruction sensitivity (p-MRR
$\approx$ 0).  We propose \method{}, a training-free decision-layer
reranker that decomposes an instruction into three scoring
views---contextual, positive-refinement, and exclusion---and resolves
competing signals at the document level.  The key insight is that
negative-channel similarity conflates true constraint violations with
background relevance; \method{} therefore estimates a query-adaptive
residual exclusion boundary from the candidate-set score distribution
and penalizes only residual exclusion evidence above this boundary,
while gating the positive reward with the same boundary to prevent
over-boosting documents that merely share background with the exclusion
topic.  Label-free query-local calibration derives the reward and
penalty scales from each query's candidate scores at inference time,
without retriever updates, relevance labels, or benchmark-specific
coefficient search.  On FollowIR, \method{} achieves the strongest
average instruction sensitivity (+14.9 p-MRR) and retrieval quality
(29.0 Score) among compared inference-time methods, outperforming the
strongest rewriting baseline (INF-X Full) by +11.5 p-MRR and +2.0
Score.  The same formulation transfers to NegConstraint without
dataset-specific retraining, and the cross-retriever evaluation
confirms portability across frozen retrievers with different embedding
spaces and score scales.
\end{abstract}

% ============================================================
% 1. INTRODUCTION
% ============================================================
\section{Introduction}
\label{sec:introduction}

Information retrieval increasingly needs to answer more than a topical
query.  Users may ask for documents that satisfy a particular condition,
avoid a confounding population, emphasize a desired aspect, or conform
to a specified relevance criterion.  In these settings,
\emph{instruction-following retrieval} (IF-IR) must rank documents by
both their topical relation to the query and their consistency with the
user's instruction.  This is a stricter requirement than conventional
semantic matching: two documents can discuss the same topic, yet only
one can satisfy the condition that makes it useful to the user
\cite{karpukhin2020dpr}.  More broadly, instruction tuning has made
language models responsive to natural-language directives
\cite{instructgpt2022,flan2022,t02022}; IF-IR requires that same
control to determine document relevance rather than generated text.

Recent benchmarks make this distinction measurable.  FollowIR
\cite{weller-etal-2025-followir} pairs a query with detailed relevance instructions
and evaluates whether a retriever changes its ranking when the
instruction changes.  Its results expose a persistent failure mode:
models that are strong at ordinary retrieval often continue to favor
documents with the right topic or keywords while underweighting the
fine-grained criteria that determine relevance.  The problem is not
limited to a particular linguistic form.  Positive requirements,
contextual qualifications, and explicit exclusions can all change which
documents should be preferred within the same topical candidate set.

Existing methods either adapt the retriever with instruction supervision
or intervene at inference time.  Promptriever \cite{weller2024promptriever},
InF-IR \cite{zhuang2025infir}, and Dual-View Training \cite{dualview2026}
adapt instruction-sensitive retrievers using additional training data and
optimization.  Training-free query rewriting, such as HyDE
\cite{gao2023hyde}, Query2Doc \cite{wang2023query2doc}, DeepRetrieval
\cite{li2025deepretrieval}, and RAG-Fusion \cite{raudaschl2024ragfusion},
instead retrieves with enhanced query representations.  It improves
semantic coverage, but supplies no document-level rule for resolving
competing instruction signals.

Explicit exclusion illustrates why this decision is difficult.  A
document that violates an exclusion is often not irrelevant: it can
share the query's topic, entities, and vocabulary with a desirable
document.  Penalizing it solely for matching an exclusion description is
also unreliable, because the same topical background can induce high
negative-channel similarity.  A more structured inference-time approach
makes this explicit at the representation layer: DEO \cite{deo2026} decomposes text queries
and optimizes their embeddings, while Negation Matters
\cite{pokhrel2026negation} decomposes visual-language queries and applies
embedding-space repulsion with context anchoring.  These methods show
that frozen backbones can be made negation-aware, but resolve competing
signals by altering query directions or similarities in the embedding
space.  We instead ask whether a frozen retriever's candidate set already
provides sufficient local evidence to make the decision at the document
level.

We propose \method{}, a training-free decision-layer reranker for IF-IR.  It
uses instruction decomposition to construct three auxiliary scoring views,
rather than replacing the original query with a single rewritten
representation: a base query ($\qbase$) that
preserves the full query and instruction, a positive refinement query
($\qpos$) that exposes affirmative evidence, and a negative exclusion
query ($\qminus$) that exposes evidence for suppression.  The views are
not a taxonomy of instruction semantics; together, they let the reranker
retain context while separately assessing helpful and potentially
conflicting signals.  Crucially, \method{} estimates the amount of
negative-channel score expected from background relevance, penalizes
only residual exclusion evidence above a query-adaptive boundary, and
uses that same boundary to gate the positive reward.  Label-free
candidate-set calibration derives the reward and penalty scales at
inference time, without retriever updates, relevance labels, or
benchmark-specific coefficient search.

We evaluate the resulting decision-layer formulation on complementary
settings.  FollowIR tests whether rankings respond to changed
instructions while preserving retrieval quality, whereas NegConstraint
\cite{xu2025logical} tests transfer to retrieval queries with explicit
logical exclusions.  Across the compared FollowIR methods, \method{}
achieves the strongest average instruction sensitivity and retrieval
quality, and the same formulation is evaluated on NegConstraint without
dataset-specific retraining.  A remaining limitation is \textbf{semantic
entanglement}---strong overlap between the base and exclusion channels---
which can make residual exclusion evidence difficult to isolate.  In
summary, our contributions are: \textbf{(i)} a
query-adaptive residual exclusion boundary and safety-gated score
composition for document-level instruction resolution; \textbf{(ii)} a
label-free query-local calibration scheme for adapting this decision
layer to candidate-set score scales; and \textbf{(iii)} an empirical
evaluation spanning changed-instruction retrieval and explicit
negative-constraint retrieval.

% ============================================================
% 2. RELATED WORK
% ============================================================
\section{Related Work}
\label{sec:related-work}

\subsection{Supervised Instruction-Following Retriever Adaptation}

Instruction-following retriever adaptation is a primary route to IF-IR.
Instruction-tuned encoders such as INSTRUCTOR \cite{instructor2022} and
TART \cite{tart2022} train embedding models with task descriptions.
More directly, Promptriever \cite{weller2024promptriever} constructs
instance-level instruction data and fine-tunes a language-model retriever
to be promptable at retrieval time; InF-IR \cite{zhuang2025infir}
likewise uses instruction supervision to train a retriever.  Dual-View
Training \cite{dualview2026} is a training-based extension that
synthesizes polarity-reversed views of a document pair.  FollowIR
\cite{weller-etal-2025-followir} formalizes the resulting requirement
through original/changed instruction pairs and p-MRR.  Together, these
works establish that instruction sensitivity can be learned, but rely on
additional data, optimization, and retriever updates.

Related LLM embedding approaches, including E5-Mistral and GritLM,
use instructions to condition general representation learning and are
typically assessed on broad embedding and retrieval suites such as MTEB
and BEIR \cite{e5mistral2023,gritlm2024,mteb2022,beir2021}.  Their
instructions usually identify a task or format, whereas IF-IR requires
the relevance definition itself to change from one query instance to the
next.

\subsection{Training-Free Instruction Resolution}

LLM-based query rewriting improves retrieval by making the user's
information need more explicit before matching.  HyDE
\cite{gao2023hyde} retrieves with hypothetical documents, Query2Doc
\cite{wang2023query2doc} enriches queries with pseudo-documents, and
RAG-Fusion \cite{raudaschl2024ragfusion} combines multiple generated
query variants.  More recent rewriting systems learn the rewriting
policy or optimize retrieval-oriented generation, including RAG-QR
\cite{ma2023ragqr} and DeepRetrieval
\cite{li2025deepretrieval}.  These methods primarily improve recall or
semantic coverage by retrieving with enhanced query representations.
Even when multiple rewrites are generated, the variants usually remain
positive descriptions of relevance rather than independent
document-level evidence for what should be suppressed.

Exclusion constraints are a particularly challenging form of
instruction following.  In exclusion queries, the hardest negatives are
not irrelevant documents but topically relevant documents that violate
the constraint.  Such documents may match the affirmative intent and
share substantial background with the query, yet violate the user's
``not'', ``without'', or ``exclude'' condition.  This makes positive
query enhancement alone less suited to the problem: it can strengthen
topical relevance without identifying which relevant-looking documents
should be suppressed.

Training-free decomposition methods make the missing exclusion signal
explicit, but differ in where they resolve its conflict with positive
relevance.  DEO \cite{deo2026} decomposes text queries into positive and
negative components and optimizes the query embedding.  In
vision-language retrieval, Negation Matters \cite{pokhrel2026negation}
similarly decomposes affirmative and negated content, then applies
embedding-space repulsion and context anchoring over a frozen CLIP
backbone.  These methods demonstrate that training-free negation
handling is possible, but their corrections are expressed as modified
query directions or representation-space similarities.  \method{}
instead keeps the retriever and its embeddings fixed, resolving
competing context, positive, and negative evidence at the candidate-set
decision layer through background-leakage estimation and query-local
calibration.

\subsection{Constraint-Aware Retrieval}

One line of work addresses negative constraints and complex logical
queries by making constraint structure explicit.  Neural-symbolic
retrieval methods such as NS-IR \cite{xu2025logical} convert
natural-language queries and documents into first-order logic forms and
enforce logical consistency through reranking objectives such as logic
alignment and connective constraints.  OrLog \cite{hoveyda2026orlog}
instead decomposes complex queries into atomic predicates, estimates
predicate-level plausibility with LLMs, and applies probabilistic
reasoning to infer query satisfaction.  These routes offer
interpretable ways to represent logical constraints, but they introduce
a different dependency: natural-language retrieval must be converted
into logical or predicate-level representations, or coupled with
symbolic/probabilistic reasoning objectives.  Our method stays in the
dense score space and does not require a logical parser, symbolic
inference module, or probabilistic reasoning engine.

\subsection{Query-Local Score Calibration and Test-Time Adaptation}

The calibration component of \method{} is related to post-hoc
calibration and local similarity correction.  Platt scaling and
temperature scaling transform model outputs after training while
keeping model parameters fixed
\cite{platt1999probabilistic,guo2017temperature}, but are typically
estimated from labeled validation data for confidence calibration.
Local scaling methods such as CSLS \cite{conneau2018wordtranslation}
show a complementary principle: raw similarities can contain local
bias, and neighborhood statistics can improve comparability across
candidates.  Our setting differs from both lines because it requires
label-free calibration of ranking scores for each query.

Information retrieval also provides a direct precedent for using
unlabeled query-local evidence.  Query performance prediction estimates
the expected reliability of a ranking without relevance judgments, and
recent studies show that neural rankers exhibit query-specific
behaviors that are difficult to capture with traditional predictors
\cite{faggioli2023query}.  Pseudo-relevance feedback similarly uses the
top-ranked set as local evidence, including in dense retrieval settings
where feedback representations are extracted from first-pass results
\cite{wang2021pseudo}.  \method{} adopts the same broad test-time
principle but uses the candidate set differently: it does not predict
query difficulty or expand the query, but estimates the score scales of
the base, positive, and negative channels.

Recent inference-only calibration work further shows that unlabeled
test batches can be used to estimate and correct score biases without
supervised validation labels \cite{zhou2024batch}.  In contrast,
\method{} does not tune prompts, update model parameters, or learn
global calibration constants.  It keeps the retriever fixed and uses
candidate-set statistics only to calibrate decision-layer reward and
penalty scales.

% ============================================================
% 3. METHOD
% ============================================================
\section{Method: DeIR-Dual}
\label{sec:method}

% Section narrative: We first formalize why instruction-following
% retrieval needs independent channels, then introduce triple-query
% decomposition, present the query-adaptive exclusion boundary, and
% finally describe the label-free inference-time calibration that
% sets the reward and penalty scales.

\subsection{Problem Formulation}
\label{sec:problem-formulation}

We consider two-stage inference with a frozen dense retriever.  Given a
query $q$ and instruction $I$, the first stage retrieves a top-$K$
candidate set $\mathcal{C}_q$ with the full-context view $\qbase$.
\method{} then reorders only $\mathcal{C}_q$ according to topical
relevance and the constraints expressed in $I$, without retriever
updates or labeled calibration data.  This separation preserves
first-stage recall while concentrating constraint resolution on the
documents that compete at the top of the ranking.  The difficulty is
not simply to make the instruction more explicit: a document that
violates an exclusion can still be highly relevant to the query topic.
Its negative-channel score therefore mixes constraint-violation evidence
with similarity caused by shared background.

\method{} resolves this ambiguity at the decision layer.  It retains a
frozen retriever and separates the available evidence into contextual,
affirmative, and exclusion views.  It then estimates how much
negative-channel similarity is explained by background relevance and
uses the residual only for candidate-level decisions.

\subsection{Triple-Query Scoring Views}
\label{sec:dual-query-generation}

An LLM decomposer constructs three short retrieval-style queries from
$(q,I)$: $\qbase$, containing the original query and full instruction;
$\qpos$, retaining affirmative constraints; and $\qminus$, containing
only excluded content.  In particular, $\qminus$ names the excluded
content rather than retaining a negation operator (e.g., ``without
X'' becomes ``X'').  These are operational scoring views, not a
taxonomy of instruction semantics.  The base view retains context that
can be lost in an affirmative/exclusion split, while the other two
views isolate evidence used by the decision layer.  If no exclusion is
present, the decomposer returns $\qminus=\textnormal{[NONE]}$ and the
negative branch is skipped.

Table~\ref{tab:decomposition-example} illustrates the decomposition.
The exclusion view is deliberately a retrievable description of the
out-of-scope content: high similarity to it is evidence to be assessed
against topical background in the next stage, rather than an instruction
executed directly in the embedding space.

\begin{table}[t]
\centering
\caption{Illustrative operational decomposition.  $\qbase$ preserves
the complete request; $\qpos$ states what should qualify as relevant;
and $\qminus$ describes only the content to exclude.}
\label{tab:decomposition-example}
\small
\setlength{\tabcolsep}{3pt}
\renewcommand{\arraystretch}{1.12}
\begin{tabular}{p{0.19\linewidth} p{0.72\linewidth}}
\toprule
\textbf{View} & \textbf{Text} \\
\midrule
Input & \emph{What is the ongoing status of the Three Gorges Project?}
Relevant documents report its completion date, total cost, or electrical
output; social, political, and ecological impacts are out of scope. \\
$\qbase$ & What is the ongoing status of the Three Gorges Project?
Relevant documents report its completion date, total cost, or electrical
output; social, political, and ecological impacts are out of scope. \\
$\qpos$ & Status of the Three Gorges Project: completion date, total
cost, and electrical output. \\
$\qminus$ & Social, political, and ecological impacts of the Three
Gorges Project. \\
\bottomrule
\end{tabular}
\end{table}

Let $f_q$ and $f_d$ denote the query and document encoders of the
unchanged retriever.  For each $d\in\mathcal{C}_q$, we compute
\begin{equation}
\begin{aligned}
 \sbase(d)&=\cos(f_q(\qbase),f_d(d)), &
 \spos(d)&=\cos(f_q(\qpos),f_d(d)),\\
 \sneg(d)&=\cos(f_q(\qminus),f_d(d)).
\end{aligned}
\label{eq:channel-scores}
\end{equation}
Document embeddings are computed once and reused for all views.

\subsection{Query-Adaptive Exclusion Boundary}
\label{sec:scoring-formula}

The base and positive channels provide topical relevance and
instructional refinement.  The negative channel is more delicate: a
high $\sneg(d)$ can indicate a true constraint violation, but it can
also arise because relevant and excluded documents share entities,
topic, or background context.  \method{} therefore treats exclusion as
a \textbf{residual boundary estimation} problem.  Negative-channel
similarity is decomposed into background leakage and residual exclusion
evidence.  We estimate the background component from the current
query's candidate distribution, subtract it from the negative-channel
score, and penalize residual evidence only as it approaches and exceeds
a robust query-specific boundary.

\subsubsection{Background Leakage Estimation}

Our treatment follows the same broad intuition as local scaling methods
for nearest-neighbor retrieval: raw similarity can contain a query-local
bias that should be removed before comparison
\cite{conneau2018wordtranslation}.  In our setting, this bias is
background-induced similarity in the negative channel.  We use a simple
query-local linear leakage model to estimate the part of the
negative-channel score that is expected from background relevance.  Let
$\mu_b,\sigma_b$ be the mean and standard deviation of
$\sbase$ over the candidate set, and let $\mu_n,\sigma_n$ be the
corresponding statistics for $\sneg$.  We first standardize the base
score,
%
\begin{equation}
  z_b(d) =
    \frac{\sbase(d) - \mu_b}{\sigma_b},
  \label{eq:base-zscore}
\end{equation}
%
and measure the query-level overlap between the base and exclusion
channels as
%
\begin{equation}
  c_q = \cos(f_q(\qbase), f_q(\qminus)).
  \label{eq:query-overlap}
\end{equation}
%
We use a query-local affine baseline for the negative score:
%
\begin{equation}
  \widehat{S}^{\text{bg}}_{\text{neg}}(d)
    = \mu_n + \sigma_n\, c_q\, z_b(d).
  \label{eq:bg-estimate}
\end{equation}
%
This baseline transfers base-score variation into the negative-channel
scale according to base/exclusion semantic overlap.  It is a
query-local control for score co-variation, rather than a relevance
model in its own right.

\subsubsection{Residual Exclusion Boundary}

We define the residual negative score as the excess negative-channel
evidence beyond the background estimate:
%
\begin{equation}
  R_{\text{neg}}(d)
    = \sneg(d) - \widehat{S}^{\text{bg}}_{\text{neg}}(d).
  \label{eq:residual}
\end{equation}
%
Positive residuals indicate that the document matches the exclusion
query more strongly than expected from background relevance alone.  We
define a robust query-adaptive residual margin:
%
\begin{equation}
  a_q = \operatorname{MAD}(\mathcal{R}_q),\qquad
  m_q = \lambda a_q,\quad
  \mathcal{R}_q=\{R_{\text{neg}}(d):d\in\mathcal{C}_q\},
  \label{eq:mad-margin}
\end{equation}
%
where $\mathcal{C}_q$ denotes the retrieved candidate set.  Residuals
above $m_q$ provide stronger evidence for exclusion.  We use
Softplus as a smooth approximation to a hinge penalty:
%
\begin{equation}
  \text{penalty}(d)
    = \alpha_q\,\operatorname{Softplus}\!\left(R_{\text{neg}}(d)-m_q\right).
  \label{eq:residual-penalty}
\end{equation}
%
The penalty grows smoothly as residual exclusion evidence approaches
and exceeds the boundary, while the residual construction prevents the
negative channel from penalizing documents merely because they share
topical background with the exclusion query.  The scale $\alpha_q$ is
derived by score calibration in \S\ref{sec:parameter-derivation}.

\subsubsection{Safety-Gated Positive Reward}

The positive channel can improve recall for documents satisfying the
instruction, but a raw positive reward can also amplify documents that
violate the exclusion constraint.  We therefore gate the positive
reward by residual evidence relative to the background expectation:
%
\begin{equation}
  g_q(d)
    = 1 - \sigma\!\left(
      \kappa \frac{R_{\text{neg}}(d)}{a_q}
    \right).
  \label{eq:safety}
\end{equation}
%
The penalty margin and the gate have distinct roles: $m_q$ determines
when an extra exclusion penalty becomes active, whereas the gate begins
to reduce positive reward once the residual exceeds the background
expectation.  The fixed shape parameter $\kappa$ controls the sharpness
of this transition.

\textbf{Final score.}
The final decision-layer score combines base relevance, safety-gated
positive refinement, and residual exclusion penalty:
%
  \begin{equation}
  \sfinal(d)
    = \sbase(d)
    + \beta_q\,\spos(d)\,g_q(d)
    - \alpha_q\,\operatorname{Softplus}\!\left(R_{\text{neg}}(d)-m_q\right).
  \label{eq:final}
\end{equation}
%
This nonlinear score composition implements a query-adaptive exclusion
boundary at ranking time without modifying the dense retriever.  The
parameters $\alpha_q$ and $\beta_q$ are derived from the score
distributions of the current query, while $\lambda$ and $\kappa$
are fixed shape parameters controlling boundary offset and gate
sharpness.  When an instruction has no exclusion, we omit the negative
branch and rank by $\sbase(d)+\beta_q\spos(d)$.  Numerical safeguards
and missing-view handling are deferred to Appendix~\ref{app:implementation}.

\begin{figure*}[t]
  \centering
  \setlength{\tabcolsep}{4pt}
  \renewcommand{\arraystretch}{1.18}
  \resizebox{0.95\linewidth}{!}{\fbox{\small
  \begin{tabular}{c@{\;\(\longrightarrow\)\;}c@{\;\(\longrightarrow\)\;}c@{\;\(\longrightarrow\)\;}c}
  \parbox[c][1.25cm][c]{2.15cm}{\centering Query $+$ instruction\\[2pt]
    \colorbox{blue!10}{\strut $\qbase$}\;\colorbox{green!10}{\strut $\qpos$}\;\colorbox{red!10}{\strut $\qminus$}}
  & \parbox[c][1.25cm][c]{1.65cm}{\centering Frozen retriever\\[2pt]
    shared document embeddings}
  & \parbox[c][1.25cm][c]{3.05cm}{\centering Candidate-set decision layer\\[2pt]
    $\widehat{s}^{\mathrm{bg}}_n \rightarrow r_q \rightarrow m_q$\\
    residual penalty $+$ safety gate}
  & \parbox[c][1.25cm][c]{2.1cm}{\centering Query-local calibration\\[2pt]
    $\alpha_q,\beta_q$\\ final reranking}
  \end{tabular}}}
  \caption{\method{} uses $\qbase$ for first-stage retrieval and keeps
    the retriever frozen.  Three scoring views provide contextual,
    affirmative, and exclusion evidence for the same top-$K$ candidate
    set.  The decision layer removes background-induced negative
    similarity, separately forms a residual penalty and a safety gate,
    and calibrates their scales from unlabeled candidate-set scores.}
  \label{fig:pipeline}
\end{figure*}

% ============================================================
% 3.4 INFERENCE-TIME SCORE CALIBRATION
% ============================================================
\subsection{Inference-Time Score Calibration}
\label{sec:parameter-derivation}

The scoring formula in Eq.~\ref{eq:final} contains two query-level
calibration factors: $\alpha_q$ for the residual exclusion penalty and
$\beta_q$ for the positive refinement reward.  Rather than
grid-searching these values on labeled queries, we derive them at
reranking time from the score distributions of the current candidate
set.  This converts parameter selection into query-conditioned score
calibration: the correction terms are placed on a magnitude comparable
to the base retriever without changing the retriever or using relevance
labels.  The goal is not to fit a benchmark-specific optimum, but to
make the exclusion boundary and positive reward portable across
retrievers with different score scales.

\subsubsection{Candidate-Set Score Calibration}
\label{sec:scale-alignment}

The base, positive, and negative channels are produced by the same dense
retriever but have different score distributions because they encode
different query intents.  We therefore treat $\alpha_q$ and $\beta_q$
as post-hoc scale factors for decision-layer reranking.  Unlike
validation-set calibration methods such as Platt scaling and
temperature scaling
\cite{platt1999probabilistic,guo2017temperature}, our calibration is
label-free, query-local, and operates on ranking scores.  The only
inputs are the channel scores over the current candidate set
$\mathcal{C}_q$.

Let the residual excess over the boundary be
%
\begin{equation}
  o_q(d) = R_{\text{neg}}(d) - m_q,
  \label{eq:residual-excess}
\end{equation}
%
and define the residual-exceeding subset as
%
\begin{equation}
  \mathcal{A}_q =
  \{d \in \mathcal{C}_q : o_q(d) > 0\}.
  \label{eq:exceeding-set}
\end{equation}
%
These are the documents whose negative residual exceeds the robust
boundary and therefore carries residual exclusion evidence.  We write
the corresponding penalty magnitude as
%
\begin{equation}
  P_q(d) = \operatorname{Softplus}(o_q(d)).
  \label{eq:penalty-magnitude}
\end{equation}
%
We derive $\alpha_q$ and $\beta_q$ by active-region scale matching.
The penalty is matched to base scores on the penalty-active set:
%
\begin{equation}
  \alpha_q =
    \frac{
      \operatorname*{Avg}_{d\in\mathcal{A}_q} \sbase(d)
    }{
      \operatorname*{Avg}_{d\in\mathcal{A}_q} P_q(d)
    }.
  \label{eq:alpha}
\end{equation}
%
For the positive channel, we use the complementary safe subset
%
\begin{equation}
  \mathcal{B}_q =
  \{d \in \mathcal{C}_q : o_q(d) \leq 0\}.
  \label{eq:safe-set}
\end{equation}
%
The reward scale is computed from the raw positive channel on the safe
set.  Keeping $g_q(d)$ out of this calibration prevents scale alignment
from compensating for the safety gate:
%
\begin{equation}
  \beta_q =
    \frac{
      \max_{d\in\mathcal{B}_q} \sbase(d)
    }{
      \operatorname*{Avg}_{d\in\mathcal{B}_q} \spos(d)
    }.
  \label{eq:beta}
\end{equation}
%
Numerical safeguards for empty active sets and near-zero denominators
are fixed before evaluation and described in Appendix~\ref{app:implementation}.
Crucially, $\alpha_q$ and $\beta_q$ use only the unlabeled score
distribution of the current candidate set; no relevance labels,
validation queries, or benchmark-level coefficient search enter
Eqs.~\ref{eq:alpha}--\ref{eq:beta}.  In contrast, $\lambda$ and
$\kappa$ are fixed shape constants for the residual margin and gate
sharpness, respectively; their values and sensitivity are reported in
Appendix~\ref{app:implementation}.
% ============================================================
% 5. EXPERIMENTS
% ============================================================
\section{Experiments}
\label{sec:experiments}

\subsection{Experimental Setup}
\label{sec:evaluation-overview}

We evaluate \method{} as an inference-time reranker over the top-$K$
candidate set produced by a frozen base retriever.  Every comparison uses
the same candidate set for its base model; the three query views share the
retriever's query template, and document embeddings are computed once.
The decomposer, prompting configuration, candidate depth, and fixed shape
constants $\lambda,\kappa$ are held fixed within each evaluation protocol;
only $\alpha_q,\beta_q$ are derived from the current unlabeled candidate
set.  Full configuration details are in Appendix~\ref{app:implementation}.

FollowIR \cite{weller-etal-2025-followir} is our primary benchmark for
changed-instruction retrieval on Robust04, News21, and Core17.  We report
its instruction-sensitivity p-MRR together with MAP/nDCG and the
benchmark's aggregate Score.  NegConstraint \cite{xu2025logical} provides
a complementary transfer test in which exclusion is expressed directly as
a retrieval constraint; there we report MAP and nDCG@10.  Unless otherwise
noted, methods are used without retriever fine-tuning, relevance labels,
or benchmark-level coefficient search.

\subsection{Main Results}
\label{sec:main-results}

\subsubsection{Instruction Following on FollowIR}

Table~\ref{tab:followir-main} reports representative results on FollowIR.
It separates the frozen base retriever, a supervised instruction-adaptation
reference (Promptriever), and representative inference-time methods.  The
direct comparison is therefore among the inference-time methods, whereas
Promptriever indicates the performance of the data-construction-and-
fine-tuning route.  The complete comparison is deferred to
Appendix~\ref{app:followir-full}.

\begin{table*}[!t]
\centering
\small
\begin{tabular}{l|cc|cc|cc|cc}
\toprule
& \multicolumn{2}{c|}{Core17} & \multicolumn{2}{c|}{Robust04} & \multicolumn{2}{c|}{News21} & \multicolumn{2}{c}{Average} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7} \cmidrule(lr){8-9}
\textbf{Method} & MAP & p-MRR & MAP & p-MRR & nDCG & p-MRR & Score & p-MRR \\
\midrule
RepLLaMA (base) & 20.6 & +1.3 & 24.0 & -8.9 & 24.5 & -1.8 & 23.0 & -3.1 \\
Promptriever$^{\dagger}$ & 21.6 & +15.4 & 28.3 & +11.7 & 28.5 & +6.4 & 26.1 & +11.2 \\
\midrule
HyDE & 23.6 & +6.5 & 24.9 & -2.9 & 21.4 & +0.7 & 23.4 & +1.4 \\
Query2Doc & 25.9 & +8.0 & 27.9 & -7.5 & 24.9 & -3.8 & 26.2 & -1.1 \\
INF-X Full & 27.0 & +7.2 & 32.1 & -1.4 & 22.3 & +4.4 & 27.0 & +3.4 \\
\midrule
DeIR-Dual & \textbf{26.0} & \textbf{+17.6} & \textbf{28.6} & \textbf{+8.3} & \textbf{32.3} & \textbf{+18.7} & \textbf{29.0} & \textbf{+14.9} \\
\bottomrule
\end{tabular}
\caption{Representative FollowIR comparison.  Promptriever$^{\dagger}$
  is a supervised instruction-trained reference; the other non-base
  baselines are inference-time methods.  All metrics are percentages
  ($\times 100$); the complete baseline table is in the appendix.}
\label{tab:followir-main}
\end{table*}

\subsubsection{Portability Across Frozen Retrievers}
\label{sec:cross-retriever-main}

Table~\ref{tab:cross-retriever} evaluates \method{} across frozen
retrievers with different score scales and embedding spaces.  For each
retriever, reranking is performed on the same top-$K$ candidate set
returned by the base model.  The same shape constants are used for all
retrievers, while $\alpha_q$ and $\beta_q$ are derived from each
query's candidate scores.  This setting tests whether query-local
calibration is tied to a single encoder or can be applied as a portable
decision-layer reranker over different frozen retrievers.

\begin{table*}[!t]
\centering
\small
\begin{tabular}{lcccc}
\toprule
& \multicolumn{2}{c}{Base Retriever} &
\multicolumn{2}{c}{\method{}} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5}
\textbf{Retriever} & \textbf{Score} & \textbf{p-MRR}
& \textbf{Score} & \textbf{p-MRR} \\
\midrule
BM25 & 15.2 & $+$1.1 & 17.0 & 17.5 \\
E5-Mistral & 38.4 & $-$2.8 & 26.3 & 16.0 \\
BGE & 31.2 & $-$2.2 & 21.1 & 19.5 \\
RepLLaMA & 37.4 & $-$3.3 & 25.2 & 20.3 \\
\bottomrule
\end{tabular}
\caption{Cross-retriever results on FollowIR.  Every row uses the same
decomposer, candidate depth, and fixed $\lambda,\kappa$; only
$\alpha_q,\beta_q$ are derived from the current candidate set.}
\label{tab:cross-retriever}
\end{table*}

\subsubsection{Transfer to Negation-Sensitive Benchmarks}
\label{sec:negation-benchmarks}

Table~\ref{tab:negation-transfer} evaluates transfer to NegConstraint,
a text-retrieval benchmark where exclusion is expressed directly as a
retrieval constraint.  Following the NegConstraint protocol
\cite{xu2025logical}, we report the total MAP and nDCG@10 in the main
text and defer formulation-level results to Appendix~\ref{app:negconstraint}.
We also include DEO \cite{deo2026}, a close training-free
negation-aware retrieval baseline.  The method is transferred without
dataset-specific retraining or coefficient search: the same scoring
mechanism and fixed shape constants are used, while $\alpha_q$ and
$\beta_q$ are derived from unlabeled candidate-set scores.  This setting
tests whether the residual boundary mechanism remains useful beyond
FollowIR's changed-instruction protocol.  The formulation-level breakdown
is reported in Appendix~\ref{app:negconstraint}.

\begin{table*}[!t]
\centering
\small
\begin{tabular}{lcc}
\toprule
\textbf{Method} & \textbf{MAP} & \textbf{nDCG@10} \\
\midrule
Base retriever & 70.8 & 77.7 \\
+ Positive channel & 76.2 & 81.6 \\
+ Raw negative penalty & 78.5 & 83.5 \\
DEO (BGE-M3)$^{\dagger}$ & 73.8 & 79.5 \\
\method{} & 75.0 & 80.4 \\
\bottomrule
\end{tabular}
\caption{Transfer to NegConstraint \cite{xu2025logical}.  DEO$^{\dagger}$
  is the reported BGE-M3 result from its arXiv preprint \cite{deo2026};
  it is an embedding-space reference rather than a matched reranking
  comparison.  All scores are percentages ($\times 100$).}
\label{tab:negation-transfer}
\end{table*}

\subsection{Ablation Study}
\label{sec:mechanism-analysis}

Table~\ref{tab:ablation-main} tests two competing explanations for the
main result.  The first block asks whether positive rewriting or a raw
negative penalty is sufficient; the second asks whether the score scales
can be fixed or calibrated globally.  We report the aggregate FollowIR
metrics here and defer per-collection breakdowns and secondary variants
to Appendix~\ref{app:additional-ablations}.

\begin{table}[t]
\centering
\small
\begin{tabular}{lcc}
\toprule
\textbf{Variant} & \textbf{Avg. Score} & \textbf{Avg. p-MRR} \\
\midrule
\multicolumn{3}{l}{\emph{Exclusion mechanism}} \\
Base retriever & 23.81 & $-$3.35 \\
+ Positive channel & 25.04 & 19.74 \\
+ Raw negative penalty & 26.61 & 1.14 \\
+ Residual boundary & 26.86 & 5.29 \\
Full \method{} & 25.33 & \textbf{20.14} \\
\midrule
\multicolumn{3}{l}{\emph{Score calibration}} \\
Fixed $\alpha,\beta$ & 25.24 & 18.53 \\
Global calibration & 25.14 & 19.90 \\
Query-local calibration & 25.23 & 20.31 \\
Grid-search oracle$^{\ddagger}$ & 24.21 & \textbf{22.29} \\
\bottomrule
\end{tabular}
\caption{Compact FollowIR ablation.  The first block contrasts competing
exclusion mechanisms; the second changes only the selection of
$\alpha,\beta$.  $^{\ddagger}$Oracle tuning is an analysis upper bound,
not a deployable method.  All metrics are percentages ($\times 100$).}
\label{tab:ablation-main}
\end{table}

We fix $\lambda$ and $\kappa$ across all retrievers and datasets.
Sensitivity, candidate-depth, efficiency, and secondary-design analyses
are reported in Appendix~\ref{app:additional-ablations}.

% ============================================================
% 7. CONCLUSION AND LIMITATIONS
% ============================================================
\section{Conclusion and Limitations}
\label{sec:conclusion}

This paper presented \method{}, a frozen-retriever decision-layer
reranking method for text retrieval under exclusion-sensitive
instructions.  The central idea is to avoid treating negative-query
similarity as uniformly harmful.  \method{} first uses triple-query
decomposition to separate contextual relevance, positive refinement, and
exclusion evidence; it then estimates a query-adaptive residual
exclusion boundary over the candidate set and applies label-free
query-local score calibration to place residual penalties and gated
positive rewards on comparable scales.  The current FollowIR results
show improvements in instruction sensitivity and aggregate retrieval
quality over inference-time rewriting and reranking baselines.  The
cross-retriever, transfer, ablation, and appendix analyses characterize
how these effects vary across frozen retrievers, datasets, mechanism
variants, and fixed shape constants.

The scope of \method{} is bounded by several limitations.  It depends
on LLM-generated decompositions, so ambiguous or underspecified
instructions can propagate errors into the positive and negative
channels.  High-overlap cases, where desired and excluded semantics are
tightly entangled, remain difficult because residual exclusion evidence
can be hard to distinguish from background relevance.  The fixed shape
constants $\lambda$ and $\kappa$ are shared across settings, so their
portability is assessed by the appendix sensitivity analysis.
Finally, because \method{} reranks an existing candidate set rather than
retrieving new documents, its effectiveness depends on the base
retriever placing relevant documents in $\mathcal{C}_q$; missed
candidates cannot be recovered by the decision-layer scorer.

% ============================================================
% APPENDIX
% ============================================================
\appendix

\section{Complete FollowIR Comparison}
\label{app:followir-full}

Table~\ref{tab:followir-full} expands Table~\ref{tab:followir-main} to
all evaluated inference-time baselines.  Values unavailable from a
baseline report are retained as \textnormal{TBD} rather than imputed.

\begin{table*}[!t]
\centering
\small
\begin{tabular}{l|cc|cc|cc|cc}
\toprule
& \multicolumn{2}{c|}{Core17} & \multicolumn{2}{c|}{Robust04} & \multicolumn{2}{c|}{News21} & \multicolumn{2}{c}{Average} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7} \cmidrule(lr){8-9}
\textbf{Method} & MAP & p-MRR & MAP & p-MRR & nDCG & p-MRR & Score & p-MRR \\
\midrule
DeepRetrieval & 21.5 & +6.6 & 25.9 & -5.3 & 23.9 & -2.2 & 23.8 & -0.3 \\
HyDE & 23.6 & +6.5 & 24.9 & -2.9 & 21.4 & +0.7 & 23.4 & +1.4 \\
Query2Doc & 25.9 & +8.0 & 27.9 & -7.5 & 24.9 & -3.8 & 26.2 & -1.1 \\
RAG-Fusion & 21.9 & +5.4 & 21.1 & -8.1 & 21.0 & +1.8 & 21.3 & -0.3 \\
RAG-QR & 22.3 & +2.3 & 27.6 & -10.3 & 22.9 & -2.1 & 24.3 & -3.4 \\
ConvSearch-R1 & 22.3 & +3.0 & 25.9 & -2.5 & 22.0 & +0.2 & 23.4 & +0.2 \\
INF-X Full & 27.0 & +7.2 & 32.1 & -1.4 & 22.3 & +4.4 & 27.0 & +3.4 \\
BGE-Reasoner-Rewriter & 21.5 & +0.7 & \textnormal{TBD} & \textnormal{TBD} & 20.5 & +3.4 & \textnormal{TBD} & \textnormal{TBD} \\
ReFeed-8B & 18.3 & -6.6 & 23.3 & -1.2 & 22.4 & -0.4 & 21.2 & -2.7 \\
TongSearch-QR & 21.7 & +3.3 & 24.5 & -4.6 & 22.6 & +0.2 & 22.7 & -0.4 \\
\midrule
DeIR-Dual & \textbf{26.0} & \textbf{+17.6} & \textbf{28.6} & \textbf{+8.3} & \textbf{32.3} & \textbf{+18.7} & \textbf{29.0} & \textbf{+14.9} \\
\bottomrule
\end{tabular}
\caption{Complete FollowIR comparison for inference-time baselines.  All
metrics are percentages ($\times 100$).}
\label{tab:followir-full}
\end{table*}

\section{Additional Ablations and Sensitivity}
\label{app:additional-ablations}

Tables~\ref{tab:ablation-mechanism-full} and
\ref{tab:ablation-calibration-full} give the per-collection breakdowns
for the compact ablation in the main text.  Table~\ref{tab:diagnostic-ablation}
then separates secondary design choices that should not be conflated with
the two central mechanisms.

\begin{table*}[!t]
\centering
\small
\begin{tabular}{l|cc|cc|cc|cc}
\toprule
& \multicolumn{2}{c|}{Core17} & \multicolumn{2}{c|}{Robust04} & \multicolumn{2}{c|}{News21} & \multicolumn{2}{c}{Average} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7} \cmidrule(lr){8-9}
\textbf{Variant} & MAP & p-MRR & MAP & p-MRR & nDCG & p-MRR & Score & p-MRR \\
\midrule
Base retriever & 23.21 & 1.07 & 25.71 & $-$9.27 & 22.52 & $-$1.84 & 23.81 & $-$3.35 \\
+ Positive channel & 24.82 & 27.44 & 23.25 & 9.57 & 27.04 & 22.21 & 25.04 & 19.74 \\
+ Raw negative penalty & 26.56 & 0.49 & 24.47 & 1.77 & 28.79 & 1.15 & 26.61 & 1.14 \\
+ Residual boundary & 26.88 & 4.74 & 23.87 & 5.11 & 29.82 & 6.03 & 26.86 & 5.29 \\
Full \method{} & 24.70 & 28.17 & 22.88 & 9.61 & 28.41 & 22.65 & 25.33 & 20.14 \\
\bottomrule
\end{tabular}
\caption{Per-collection exclusion-mechanism ablation on FollowIR.}
\label{tab:ablation-mechanism-full}
\end{table*}

\begin{table*}[!t]
\centering
\small
\begin{tabular}{l|cc|cc|cc|cc}
\toprule
& \multicolumn{2}{c|}{Core17} & \multicolumn{2}{c|}{Robust04} & \multicolumn{2}{c|}{News21} & \multicolumn{2}{c}{Average} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7} \cmidrule(lr){8-9}
\textbf{Scale selection} & MAP & p-MRR & MAP & p-MRR & nDCG & p-MRR & Score & p-MRR \\
\midrule
Fixed $\alpha,\beta$ & 24.67 & 27.26 & 23.59 & 6.96 & 27.46 & 21.38 & 25.24 & 18.53 \\
Global calibration & 24.70 & 27.92 & 23.31 & 9.24 & 27.40 & 22.55 & 25.14 & 19.90 \\
Query-local calibration & 24.96 & 28.15 & 22.85 & 10.02 & 27.87 & 22.75 & 25.23 & 20.31 \\
Grid-search oracle & 22.17 & 31.68 & 22.77 & 10.20 & 27.70 & 25.00 & 24.21 & 22.29 \\
\bottomrule
\end{tabular}
\caption{Per-collection calibration ablation on FollowIR.}
\label{tab:ablation-calibration-full}
\end{table*}

\begin{table*}[!t]
\centering
\small
\begin{tabular}{llccc}
\toprule
\textbf{Variant} & \textbf{Change from Full \method{}} &
\textbf{Avg Score} & \textbf{Avg p-MRR} & \textbf{Runtime} \\
\midrule
Full \method{} & None & 25.3 & 20.1 & $<$1\,s \\
w/o safety gate & Remove $g_q(d)$ from positive reward & 26.6 & 1.1 & $<$1\,s \\
$\alpha_q$ only & Query-local penalty scale, fixed reward scale & 25.0 & 19.7 & $<$1\,s \\
$\beta_q$ only & Query-local reward scale, fixed penalty scale & 25.3 & 4.3 & $<$1\,s \\
Mean/std boundary & Replace median/MAD boundary with mean/std & 25.2 & 19.3 & $<$1\,s \\
Raw $s_n$ boundary & Estimate boundary on raw negative similarity & 26.9 & 5.3 & $<$1\,s \\
\bottomrule
\end{tabular}
\caption{Diagnostic ablations for secondary design choices.  The table
keeps the same frozen retriever and candidate pool as
Table~\ref{tab:ablation-main}.}
\label{tab:diagnostic-ablation}
\end{table*}

Table~\ref{tab:sensitivity-efficiency} records robustness and efficiency
analyses.  \textnormal{TBD} marks results retained for experiments that
have not yet been run; no values are imputed.

\begin{table*}[!t]
\centering
\small
\begin{tabular}{llcccc}
\toprule
\textbf{Analysis} & \textbf{Setting} & \textbf{Avg Score} &
\textbf{Avg p-MRR} & \textbf{Latency} & \textbf{Notes} \\
\midrule
$\lambda$ sensitivity & $\lambda \in \{0.5, 1.0, 1.5, 2.0, 2.5, 3.0\}$ & $25.3 \pm 0.0$ & $16.7 \pm 0.0$ & $<$1\,s & Insensitive \\
$\kappa$ sensitivity & $\kappa \in \{1, 2, 4, 8, 10, 15, 20\}$ & $25.3 \pm 0.8$ & $20.1 \pm 3.5$ & $<$1\,s & Trade-off \\
Candidate set size & $K \in \{10, 50, 100, 200\}$ & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & To be measured \\
Standard retrieval control & BEIR/TREC subset & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & To be measured \\
Failure subset & High semantic-overlap queries & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & To be analyzed \\
\bottomrule
\end{tabular}
\caption{Sensitivity, efficiency, and robustness analyses.  \textnormal{TBD}
entries are explicit placeholders for matched evaluation runs.}
\label{tab:sensitivity-efficiency}
\end{table*}

\section{Standard Retrieval Preservation}
\label{app:standard-retrieval}

This control asks whether the decision layer perturbs ordinary retrieval
when no exclusion is present.  We evaluate a representative BEIR subset
covering biomedical, financial, argumentative, scientific, and duplicate-
question retrieval.  For each backbone, we compare the frozen base query,
the same query paired with a task-neutral retrieval instruction, and
\method{} with $\qminus=\textnormal{[NONE]}$.  The latter condition uses
the same candidate-generation protocol as the neutral-instruction
condition and applies no exclusion penalty.  We report average nDCG@10
and MAP@100 over the selected BEIR tasks; these ranking metrics are
nontrivial under candidate-set reranking, unlike candidate recall at an
unchanged cutoff.

\begin{table*}[!t]
\centering
\small
\begin{tabular}{lcc|cc|cc}
\toprule
& \multicolumn{2}{c|}{Base query} &
\multicolumn{2}{c|}{+ Neutral instruction} &
\multicolumn{2}{c}{+ \method{}} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7}
\textbf{Backbone} & nDCG@10 & MAP@100 & nDCG@10 & MAP@100 & nDCG@10 & MAP@100 \\
\midrule
RepLLaMA & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} \\
E5-Mistral & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} \\
BGE & 0.613 & 0.429 & 0.613 & 0.429 & 0.601 & 0.418 \\
\bottomrule
\end{tabular}
\caption{Standard-retrieval preservation on a representative BEIR subset
(TREC-COVID, NFCorpus, FiQA-2018, ArguAna, SciFact, and Quora).  A
neutral instruction isolates input-format effects; the final condition
tests whether \method{} changes ordinary ranking when no exclusion is
active.  \textnormal{TBD} entries are to be filled from matched runs.}
\label{tab:standard-retrieval}
\end{table*}

If the neutral-instruction and \method{} columns remain close to the
base-query column, the experiment supports the claim that DeIR-Dual adds
constraint resolution without systematically degrading generic retrieval.
Any material degradation should instead be reported as a deployment
boundary and analyzed by backbone or task category.

\section{Additional NegConstraint Results}
\label{app:negconstraint}

Table~\ref{tab:negconstraint-full} provides the formulation-level
NegConstraint results corresponding to the total scores reported in
Table~\ref{tab:negation-transfer}.  DEO reports only total scores in
the referenced preprint, so unavailable formulation-level entries are
marked \textnormal{TBD}.

\begin{table*}[!t]
\centering
\small
\setlength{\tabcolsep}{3.4pt}
\begin{tabular}{lcc|cc|cc|cc}
\toprule
\multirow{2}{*}{\textbf{Method}}
& \multicolumn{2}{c|}{$A-a$}
& \multicolumn{2}{c|}{$(A-a)\cup B$}
& \multicolumn{2}{c|}{$(A-a)\cup(B-b)$}
& \multicolumn{2}{c}{Total} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7} \cmidrule(lr){8-9}
& MAP & nDCG@10 & MAP & nDCG@10 & MAP & nDCG@10 & MAP & nDCG@10 \\
\midrule
BM25 & 32.1 & 34.6 & 31.2 & 34.7 & 29.3 & 31.5 & 31.4 & 33.7 \\
Contriever & 34.8 & 36.6 & 32.1 & 33.3 & 30.9 & 32.7 & 31.8 & 35.7 \\
BGE & 37.9 & 40.5 & 34.8 & 36.8 & 33.7 & 34.8 & 36.3 & 40.8 \\
HyDE & 50.7 & 55.3 & 48.6 & 51.5 & 45.7 & 50.6 & 47.8 & 53.1 \\
InteR & 52.6 & 55.8 & 50.3 & 48.7 & 51.5 & 49.3 & 52.3 & 54.5 \\
NS-IR (GPT-4o) & 54.7 & 57.9 & 53.3 & 54.2 & 51.7 & 53.7 & 53.3 & 56.5 \\
\midrule
BGE-S & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & 67.0 & 73.7 \\
BGE-S w/ DEO & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & 73.0 & 78.0 \\
BGE-L & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & 63.0 & 71.4 \\
BGE-L w/ DEO & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & 73.3 & 78.8 \\
BGE-M3 & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & 63.7 & 72.5 \\
BGE-M3 w/ DEO & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & \textnormal{TBD} & 73.8 & 79.5 \\
\midrule
Base retriever & 74.5 & 80.6 & 62.7 & 71.4 & \textnormal{TBD} & \textnormal{TBD} & 70.8 & 77.7 \\
+ Positive channel & 81.4 & 85.5 & 64.8 & 73.2 & \textnormal{TBD} & \textnormal{TBD} & 76.2 & 81.6 \\
+ Raw negative penalty & 83.7 & 87.4 & 67.3 & 75.0 & \textnormal{TBD} & \textnormal{TBD} & 78.5 & 83.5 \\
\method{} & 80.6 & 84.4 & 62.8 & 71.6 & \textnormal{TBD} & \textnormal{TBD} & 75.0 & 80.4 \\
\bottomrule
\end{tabular}
\caption{Formulation-level NegConstraint results.  ACL reference results
  are taken from the NegConstraint paper \cite{xu2025logical}.  DEO
  results are total-only numbers reported in its arXiv preprint
  \cite{deo2026}.  All scores are percentages ($\times 100$).}
\label{tab:negconstraint-full}
\end{table*}

\section{Qualitative Analysis}
\label{app:qualitative}

\textnormal{TBD:} Add one successful decomposition and one
semantic-entanglement case.  For each case, report the original query and
instruction, the three generated views, representative candidate
documents, their channel scores, and the resulting ranking change.

\section{Implementation Details for Decomposition and Calibration}
\label{app:implementation}

The decomposer receives the original query and instruction and returns a
JSON object with an affirmative field and an exclusion field.  It is
prompted to keep the affirmative field as a fluent retrieval query, to
express exclusions as content rather than as logical negation tokens,
and to return \textnormal{[NONE]} when the instruction contains no
exclusion.  $\qbase$ concatenates the original query and full
instruction, while $\qpos$ and $\qminus$ are the two returned fields.
Malformed outputs are retried with the same format constraint; a missing
exclusion is treated as \textnormal{[NONE]}.  The exact prompt, model
identifier, decoding configuration, and generated query files are
included with the artifact.

All document representations are computed once with the base
retriever's document template, and all three query views use its query
template.  For numerical stability, standard deviations and MAD values
are lower-bounded by $10^{-6}$.  If the active or safe subset is empty,
or a calibration denominator is below this tolerance, the corresponding
scale retains its fixed fallback value.  Derived $\alpha_q$ and
$\beta_q$ are clipped to $[0.05,5.0]$.  The shape constants $\lambda$
and $\kappa$ are fixed before evaluation, shared by all datasets and
retrievers, and never selected using relevance labels or evaluation
metrics.  The released evaluation configuration records their single
reported values and the candidate-set depth for each experiment.

% ============================================================
% REFERENCES
% ============================================================
\bibliographystyle{plain}
\bibliography{../refs/references}

\end{document}
